Student Log-Data from a Randomized Evaluation of Educational Technology: A Causal Case Study

### 2020-06-15 paper.Rnw
correct posterior prob b1>0, ci

<<init,include=FALSE>>=
library(knitr)
opts_chunk$set(
echo=FALSE,results='asis',warning = FALSE,message = FALSE,cache=TRUE,error = FALSE
)
pn <- function(...) prettyNum(...,big.mark=',')
library(xtable)
library(english)

@

@@

Figure \ref{fig:psResults} shows the estimated linear function
$\hat{\tau}(r)$.
The left panel shows the CTA1 treatment effect, $Y(1)-Y(0)$ as
$r=\tau(1)$ varies; the black line shows the estimate (i.e. posterior
mean) and the red lines are random draws from the posterior
distribution, showing estimation uncertainty.
Eighty-eight percent of posterior draws of the slope of the
$\hat{\tau}(\cdot)$ function were positive, implying a posterior
probability of \Sexpr{sprintf('%1.2f',probPos)}
that students who asked for hints
with greater regularity benefited more from the CTA1 curriculum.
The right hand panel plots one posterior draw of $\eta(1)$
alongside observed outcomes $Y$.
The figure also includes estimated regression lines for the two
treatment groups.
Although, as observed in Sections \ref{sec:observational} and
\ref{sec:mediation}, $\eta(1)$ is anticorrelated with $Y$, the slope
is steeper in the control group than in the treatment group.
The distance between the two regression lines is the treatment effect,
which grows with $\eta(1)$.

@@

\section{Comparing Strategies}\label{sec:synthesis}
% The goals, assumptions, and processes of matching, mediation analysis,
% and principal stratification, when applied to implementation data from
% an EdTech RCT, each differ from each other, but also overlap in
% interesting ways.
In the case discussed here, the observational study and mediation
approaches appear to give the opposite conclusion from principal
stratification---matching and mediation suggest that excessive hint
requests hurt students' posttest scores, while principal
stratification suggests that students who request more hints
experience higher treatment effects.
This section will compare and contrast the three methods in terms of
 their necessary identification assumptions, their model fitting
 mechanics, and interpretation of their results.

@@

Like in the observational study and in mediation analysis, the
principal stratification analysis found a negative correlation
between hint requesting and posttest scores.
Those students who request more hints are those who need more help of
some form, and tend to score lower on the posttest.
Fortunately, the principal stratification analysis also suggests that these students
may benefit more from CTA1 than their peers who request fewer hints.
However, it is unclear whether their larger treatment effects are due
to their higher rate of hint requests, or to some other related characteristic.

Can these results be reconciled with the results from the previous two
sections?
Does requesting hints help or hurt?
Technically, principal stratification has nothing to say about the effect of hints---$\eta(1)$
is modeled as a student characteristic, not an intervention.
(See \citealt{jin2008principal} for a nice discussion of this point.)
Still, doesn't the correlation between treatment effects and hint
requests suggest that hints may be beneficial?
Does that contradict the results in Sections \ref{sec:observational}
and \ref{sec:mediation} that hints requests lower students' posttest
scores?
In fact, the results of the three methods can be reconciled.
For instance, it may be that requesting hints indeed lowers test
scores---a negative indirect effect---but that those students who are
likely to request more hints also tend to experience greater direct
effects.
That is, it may be that the observational and mediational results are
correct---that requesting hints lowers test scores---but that the principal stratification
results are also correct---students who request more hints tend to
experience higher treatment effects, due to other mechanisms.
(Finally, we should note that the principal stratification model was
uncertain about the sign of $b_1$---there is a decent (though not large) probability that the
the slope between $\eta(1)$ and treatment effects is negative.)

>>>>

<<init,include=FALSE>>=
library(knitr)
opts_chunk$set(
echo=FALSE,results='asis',warning = FALSE,message = FALSE,cache=TRUE,error = FALSE
)
pn <- function(...) prettyNum(...,big.mark=',')
library(xtable)
library(english)
cap <- function(x){
  x <- as.character(x)
  paste0(toupper(substr(x,1,1)),substr(x,2,nchar(x)))
}
@

@@

Figure \ref{fig:psResults} shows the estimated linear function
$\hat{\tau}(r)$.
The left panel shows the CTA1 treatment effect, $Y(1)-Y(0)$ as
$r=\tau(1)$ varies; the black line shows the estimate (i.e. posterior
mean) and the red lines are random draws from the posterior
distribution, showing estimation uncertainty.
\Sexpr{cap(english(round(probPos*100)))} percent of posterior draws of the slope of the
$\hat{\tau}(\cdot)$ function were positive, implying a posterior
probability of \Sexpr{sprintf('%1.2f',probPos)}
that students who asked for hints
with greater regularity benefited more from the CTA1 curriculum.
A 95\% highest-density credible interval for the slope parameter is
\Sexpr{printci(b1ci)}.
Therefore, the data are consistent with either a slightly negative or
a positive relationship between hint requests and treatment
effects, but the latter is more likely.

The right hand panel plots one posterior draw of $\eta(1)$
alongside observed outcomes $Y$.
The figure also includes estimated regression lines for the two
treatment groups.
Although, as observed in Sections \ref{sec:observational} and
\ref{sec:mediation}, $\eta(1)$ is anticorrelated with $Y$, the slope
is steeper in the control group than in the treatment group.
The distance between the two regression lines is the treatment effect,
which grows with $\eta(1)$.

@@

\section{Comparing Strategies}\label{sec:synthesis}
% The goals, assumptions, and processes of matching, mediation analysis,
% and principal stratification, when applied to implementation data from
% an EdTech RCT, each differ from each other, but also overlap in
% interesting ways.
In the case discussed here, the observational study and mediation
approaches appear to give the opposite conclusion from principal
stratification---matching and mediation suggest that excessive hint
requests hurt students' posttest scores, while principal
stratification suggests that students who request more hints may
experience higher treatment effects.
This section will compare and contrast the three methods in terms of
 their necessary identification assumptions, their model fitting
 mechanics, and interpretation of their results.

@@

The
principal stratification analysis also found a negative correlation
between hint requesting and posttest scores.
Those students who request more hints are those who need more help of
some form, and tend to score lower on the posttest.
Fortunately, the principal stratification analysis also suggests (with
probability \Sexpr{sprintf('%1.2f',probPos)}) that these students
may benefit more from CTA1 than their peers who request fewer hints.
However, it is unclear whether their larger treatment effects are due
to their higher rate of hint requests, or to some other related characteristic.

Can these results be reconciled with the results from the previous two
sections?
Does requesting hints help or hurt?
Technically, principal stratification has nothing to say about the effect of hints---$\eta(1)$
is modeled as a student characteristic, not an intervention.
(See \citealt{jin2008principal} for a nice discussion of this point.)
Still, doesn't the correlation between treatment effects and hint
requests suggest that hints may be beneficial?
Does that contradict the results in Sections \ref{sec:observational}
and \ref{sec:mediation} that hints requests lower students' posttest
scores?
In fact, the results of the three methods can be reconciled.
For instance, it may be that requesting hints indeed lowers test
scores---a negative indirect effect---but that those students who are
likely to request more hints also tend to experience greater direct
effects.
That is, it may be that the observational and mediational results are
correct---that requesting hints lowers test scores---but that the principal stratification
results are also correct---students who request more hints tend to
experience higher treatment effects, due to other mechanisms.
(Of course, the data are also consistent with the possibilities that
the relationship between hint requests and treatment effects is
negative, or close to zero, in which case there is less that needs to
be explained.)

### 2020-06-15 paper.tex
correct posterior prob b1>0, ci

Figure \ref{fig:psResults} shows the estimated linear function
$\hat{\tau}(r)$.
The left panel shows the CTA1 treatment effect, $Y(1)-Y(0)$ as
$r=\tau(1)$ varies; the black line shows the estimate (i.e. posterior
mean) and the red lines are random draws from the posterior
distribution, showing estimation uncertainty.
Eighty-eight percent of posterior draws of the slope of the
$\hat{\tau}(\cdot)$ function were positive, implying a posterior
probability of 0.75
that students who asked for hints
with greater regularity benefited more from the CTA1 curriculum.
The right hand panel plots one posterior draw of $\eta(1)$
alongside observed outcomes $Y$.
The figure also includes estimated regression lines for the two
treatment groups.
Although, as observed in Sections \ref{sec:observational} and
\ref{sec:mediation}, $\eta(1)$ is anticorrelated with $Y$, the slope
is steeper in the control group than in the treatment group.
The distance between the two regression lines is the treatment effect,
which grows with $\eta(1)$.

@@

Like in the observational study and in mediation analysis, the
principal stratification analysis found a negative correlation
between hint requesting and posttest scores.
Those students who request more hints are those who need more help of
some form, and tend to score lower on the posttest.
Fortunately, the principal stratification analysis also suggests that these students
may benefit more from CTA1 than their peers who request fewer hints.
However, it is unclear whether their larger treatment effects are due
to their higher rate of hint requests, or to some other related characteristic.

Can these results be reconciled with the results from the previous two
sections?
Does requesting hints help or hurt?
Technically, principal stratification has nothing to say about the effect of hints---$\eta(1)$
is modeled as a student characteristic, not an intervention.
(See \citealt{jin2008principal} for a nice discussion of this point.)
Still, doesn't the correlation between treatment effects and hint
requests suggest that hints may be beneficial?
Does that contradict the results in Sections \ref{sec:observational}
and \ref{sec:mediation} that hints requests lower students' posttest
scores?
In fact, the results of the three methods can be reconciled.
For instance, it may be that requesting hints indeed lowers test
scores---a negative indirect effect---but that those students who are
likely to request more hints also tend to experience greater direct
effects.
That is, it may be that the observational and mediational results are
correct---that requesting hints lowers test scores---but that the principal stratification
results are also correct---students who request more hints tend to
experience higher treatment effects, due to other mechanisms.
(Finally, we should note that the principal stratification model was
uncertain about the sign of $b_1$---there is a decent (though not large) probability that the
the slope between $\eta(1)$ and treatment effects is negative.)

>>>>

Figure \ref{fig:psResults} shows the estimated linear function
$\hat{\tau}(r)$.
The left panel shows the CTA1 treatment effect, $Y(1)-Y(0)$ as
$r=\tau(1)$ varies; the black line shows the estimate (i.e. posterior
mean) and the red lines are random draws from the posterior
distribution, showing estimation uncertainty.
Seventy-five percent of posterior draws of the slope of the
$\hat{\tau}(\cdot)$ function were positive, implying a posterior
probability of 0.75
that students who asked for hints
with greater regularity benefited more from the CTA1 curriculum.
A 95\% highest-density credible interval for the slope parameter is
[-0.06,0.12].
Therefore, the data are consistent with either a slightly negative or
a positive relationship between hint requests and treatment
effects, but the latter is more likely.

The right hand panel plots one posterior draw of $\eta(1)$
alongside observed outcomes $Y$.
The figure also includes estimated regression lines for the two
treatment groups.
Although, as observed in Sections \ref{sec:observational} and
\ref{sec:mediation}, $\eta(1)$ is anticorrelated with $Y$, the slope
is steeper in the control group than in the treatment group.
The distance between the two regression lines is the treatment effect,
which grows with $\eta(1)$.

@@

The
principal stratification analysis also found a negative correlation
between hint requesting and posttest scores.
Those students who request more hints are those who need more help of
some form, and tend to score lower on the posttest.
Fortunately, the principal stratification analysis also suggests (with
probability 0.75) that these students
may benefit more from CTA1 than their peers who request fewer hints.
However, it is unclear whether their larger treatment effects are due
to their higher rate of hint requests, or to some other related characteristic.

Can these results be reconciled with the results from the previous two
sections?
Does requesting hints help or hurt?
Technically, principal stratification has nothing to say about the effect of hints---$\eta(1)$
is modeled as a student characteristic, not an intervention.
(See \citealt{jin2008principal} for a nice discussion of this point.)
Still, doesn't the correlation between treatment effects and hint
requests suggest that hints may be beneficial?
Does that contradict the results in Sections \ref{sec:observational}
and \ref{sec:mediation} that hints requests lower students' posttest
scores?
In fact, the results of the three methods can be reconciled.
For instance, it may be that requesting hints indeed lowers test
scores---a negative indirect effect---but that those students who are
likely to request more hints also tend to experience greater direct
effects.
That is, it may be that the observational and mediational results are
correct---that requesting hints lowers test scores---but that the principal stratification
results are also correct---students who request more hints tend to
experience higher treatment effects, due to other mechanisms.
(Of course, the data are also consistent with the possibilities that
the relationship between hint requests and treatment effects is
negative, or close to zero, in which case there is less that needs to
be explained.)
